Newton's Method in Machine Learning

### 2025-04-09 latex/Draft/Draft.tex
saving progress

\title{Optimizing Newton's Method for Machine Learning }

\author{}
\date{}

@@

\section*{Newton's method in optimization}
Missing text goes here

@@

\section*{References}

>>>>

\title{Newton's Method in Machine Learning }

\author{Ibrahim Ahmad, Cipriano Dorbessan, Christopher Gossmann}
\date{April 2025}

@@

\section*{Proofs}
There are a few nontrivial statements that are crucial for understanding Numerical Optimization and training Neural Networks.
\subsection*{Proof Of Taylor's Theorem}
Taylor's Theorem is cruicial for doing numerical optimization. It allows us to provide an approximation of a function in the open neighbourhood of a point.

We must prove Taylor's theorem for the high dimensional case. We shall do so directly.
Let $f : \mathbb{R}^n \mapsto \mathbb{R}$. We shall observe some properties of $f$ in the region of $A$, where $A$ is an open subset of $\mathbb{R}^n$. We shall make the following assumptions: $f$ is continuous on $A$, $f$ is differentiable on $A$. (Note: We do not contain the boundary points as this allows us to handle asymptotics more neatly). 
Let us define the derivative of $f$ in the direction of $\alpha$ by writing \begin{equation}\vec{\alpha} \cdot \vec{\nabla} f(\vec{x}) = \lim_{t \to 0}\frac{f(\vec{x} + \vec{\alpha}t) - f(\vec{x})}{t}\end{equation}, under the assumption that $\vec{x} + t \vec{\alpha} \in A$. This allows us to find the derivative in any direction from the location of $x$. Let us define $\Phi(t) = f(\vec{x} + \vec{\alpha} t)$. Because $\Phi : \mathbb{R} \mapsto \mathbb{R}$
we can use Taylor's theorem on $\Phi$. This lets us write the equivalency 
\begin{equation}\Phi(t) = \Phi(0) + \Phi'(0)t + \Phi''(0)\frac{t^2}{2} + \dots\end{equation}
By substituting the definition of $\Phi(t)$ into both sides, we can write 
\begin{equation} f(\vec{x} + t \cdot \vec{\alpha}) = \Phi(0) + \Phi'(0)t + \Phi''(0)\frac{t^2}{2} + \dots\end{equation}
We also know that 
\begin{equation}\Phi(t)' = \lim_{t \to 0}\frac{f(\vec{x} + \vec{\alpha}t) - f(\vec{x})}{t} = \vec{\alpha} \cdot \vec{\nabla}f(\vec{x})\end{equation}
By the definition of $\Phi$, and $\vec{\alpha} \cdot \vec{\nabla}$.
By substituting into both sides, we get the following. For convenience, we shorten $\vec{\alpha}\cdot\vec{\nabla}$ to $\vec{\nabla}$.
\begin{equation} f(\vec{x} + t \cdot \vec{\alpha}) = f(\vec{x}) + [\vec{\nabla}f(\vec{x})]^T\vec{\alpha} t + \frac{1}{2}\vec{\alpha}^T[\vec{\nabla}^2f(\vec{x})]\vec{\alpha} \cdot t^2 + \dots\end{equation}
We shall only use first and second order methods, so we have not shown the third order term of the Taylor Series, which requires rank 3 tensor multiplications.
Thus we have proved Taylor's Theorem for functions $f : \mathbb{R}^n \mapsto \mathbb{R}$, under an open subset of $\mathbb{R}^n$.

\subsection*{Convergence Of Gradient Descent}
While we do not use Gradient Descent to train any of the models, understanding Gradient Descent and it's properties is crucial to understanding Adam, Stochastic Gradient Descent, and Newton's Method. For further reading regarding these algorithms we reccomend \cite{garrigos_2023_handbook}.

Because the Neural Network is not L-Smooth, or Convex, we must show the convergence of Gradient Descent under minimal assumptions. Let $x_k$ be the value of the descent's algorithm's value at the $k$'th iteration. We shall use the following update rule $x_{k+1} = x_k - \eta_k\vec{\nabla}f(x_k)$.

By the mean value theorem, we know that $f(\vec{x_{k+1}}) = f(\vec{x_k}) - \eta \|\vec{\nabla}f(x_k)\|^2 + o(\eta\|\vec{\nabla}f(\vec{x_k}\|^2)$. The proof of the $o$ term is complex and not useful for what we intend to do. It can be imagined as an upper bound of the remainder of the Taylor approximation $f(x_{k+1})$ after the first term. If $\eta_k$ is sufficently small, we can write $f(x_{k+1}) \leq f(x_k) - \eta_k \|\vec{\nabla}f(x_k)\|^2$. We must make the assumption that $\eta_k$ is sufficently small, as without that assumption we may overshoot the point we intend on converging to. We note, that under these assumptions $f(x_{k+1}) < f(x_k)$ if $\eta_k \|\vec{\nabla}f(x_k)\|^2$ is nonzero, and $f(x_{k+1}) = f(x_k)$ if $\eta_k \|\vec{\nabla}f(x_k)\|^2$ is zero. Thus we have shown that the function values are decreasing, and the algorithm converges at a minima or saddle point if $f$ is lower bounded.

We note that $\lim_{k \to \infty} \|\vec{\nabla}\| = 0$, as that is the state the algorithm must converge to to be considered to have converged.
We also know that $\eta_k > 0$, as if $\eta_k = 0$, then the algorithm would not converge.

\subsection*{Convergence Of Other Descent Algorithms}

We shall now study the convergence of algorithms of the following form $x_{k+1} = x_k - S_k\vec{\nabla}f(x_k)$. The constraints we place are that $S_k\vec{\nabla}f(x_k)$ is sufficiently small as to not overshoot the extrema, and that $S_k$ is a positive definite matrix. For example, Newton's method chooses $S_k = \vec{\nabla}^{-2}f(x_k)$.

Algorithms that follow this form are guarunteed to converge to a minima such that $\vec{\nabla}f(x_k) = 0$. Recall the following property of positive definite matrices, for some arbitrary vector $\vec{v}$, $\vec{v}^TS\vec{v} > 0$. In other words, when multiplying the matrix $S$ by some vector, that vector is rotated by less than $90^{\circ}$. Thus we know that $(S_k\vec{\nabla}f(x_k)) \cdot \vec{\nabla}f(x_k) > 0$. Let us now project $S_k\vec{\nabla}f(x_k)$ onto $\vec{\nabla}f(x_k)$, yielding $d_k$. Because $S_k$ is semi definite, we know that $d_k$ is nonzero and $d_k \cdot ,\vec{\nabla}f(x_k)$ or in other words, $S_k$ transforms the gradient vector such that it no longer faces the direction of steepest descent, but it still faces a direction of descent. Because the projection in the descent direction is a nonzero positive scaled version of the gradient, the descent can be considered to be similar to that of gradient descent with a different stepsize.

@@

\bibliographystyle{IEEEtran}
\bibliography{main}
